---
title: "Data Structures in R"
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float:
      smooth_scroll: true   
---



```{r echo=FALSE}
options(repos=structure(c(CRAN="https://cloud.r-project.org/")))

```

# Welcome to R for Biologists Workshop!

```{r echo=FALSE}
img1 <- magick::image_read("Bioinformatics_bootcamp.png")
img1 <- magick::image_scale(img1,"2000x2000")
plot(img1)
```

<div style="margin-bottom: 1000px;">
</div>

# Workflow

```{r, echo=FALSE}
#install.packages("magick")
#library("magick")
img <- magick::image_read("Workflow.png")
plot(img)
```

<div style="margin-bottom: 1000px;">
</div>

# Real world example of basic skills

 

```{r echo=FALSE}

library(readr)
seq1=read_file("SEQ1.txt")
seq2=read_file("SEQ2.txt")


```

Lets say we have two sequences



Sequence 1

```{r}
print(seq1)
```


& Sequence 2

```{r}
print(seq2)
```



And you want to know whether the first sequencing file is the same as the second. 

<div style="margin-bottom:500px;">
<details>
<summary>Are the 2 sequences the same?</summary>


```{r} 

seq1==seq2

```
</div>
</details>



# Comments

An important tool for writing R code is comments. These are preceded by # and are ignored by R 

```{r echo=T}
#This code will say "Hello World!"

"Hello world!"
```

Comments are a helpful tool for conveying information about your code to others.

<div style="margin-bottom:1000px;">
</div>

# 5 basic data structures in R

* Character
* Numeric
* Integer
* Logical
* Complex

## Character

Holds "string" data. For example: 

```{r}
"Hello World"
```

You can ask R to tell you what type of data you have:

```{r}
class("Hello World")
```

<div style="margin-bottom:1000px;">
</div>

## Numbers

Numbers can be integers or numerics (i.e., floats) in R. For example: 

```{r}
1

2.2

3E100

```

The default for R is to make every number a numeric

```{r}
class(1)
```

Which means that we have hidden decimal points behind the 1 (1.00 for example)

<div style="margin-bottom:1000px;">
</div>


### Specifying an integer
 
But what if we wanted specifically an integer?

```{r}
1L

class(1L)
```

<div style="margin-bottom:1000px;">
</div>

### Challenge Question 1

Is this a numeric or a character? 


```{r}
"2"
```

<details>
<summary>What is the class("2")</summary>
```{r}
class("2")
```

</details>

<div style="margin-bottom:1000px;">
</div>



## Logicals

Logicals are either True of False. For example

```{r}
TRUE

FALSE

class(TRUE)

class(FALSE)
```

...its time for another challenge question!
<div style="margin-bottom:1000px;">
</div>

### Challenge question 2

R is *case sensitive*. 

With that knowledge, what do you think the below would give us? 

<details>
<summary>What is the class(true)</summary>
```{r, error=TRUE}
class(true)
```
</details>

<div style="margin-bottom:1000px;">
</div>


## Data Structure Classes cont..

So far we have only been using class()

But we can also use a function to see the structure of any R object: 

```{r}
str(1)

str("Hello World!")

str(TRUE)
```

<div style="margin-bottom:1000px;">
</div>

### Challenge Question 3

R is an object oriented language, and it tends to put everything in some sort of class. 
What would the below tell us? 

```{r results='hide'}
str(class)
```

<details>
<summary>What is the str(class)</summary>
```{r echo = FALSE}
str(class)
```
</details>
<div style="margin-bottom:1000px;">
</div>

# Simple Data Structures Methods

## Arithmetic Methods


* Addition

```{r}
1+1

```


* Subraction

```{r}
1-1
```


* Multiplication

```{r}
2*2
```


* Division

```{r}
10/2
```

<div style="margin-bottom:1000px;">
</div>


## Equivalence Comparisons


Equivalence comparisons are a way to check if any two objects are the same

```{r}
#Does 1 equal 1?

1==1

#Does "Hello" equal "World?"

"Hello" == "World"

```


We can also invert an equivalancy comparison to check if two objects are *not* equal

```{r}
#Does 1 not equal 1?

1!=1

```

<div style="margin-bottom: 1000px;">
</div>

## Mathematical Comparisons

For numeric data types, mathematical comparisons can also be made

```{r}
#Is 1 less than 100?
1<100

#Is 2+2 greater than 2^2

2+2>2^2

```



<div style="margin-bottom: 1000px;">
</div>

### Challenge Question 4

Arithmetic in R follows PEMDAS

```{r}
4+5*3
```

vs 

```{r}
(4+5)*3
```


What would we get with the following? 

```{r results='hide'}
4+5*3>(4+5)*3
```
<details>
<summary>What would you get with 4+5x3>(4+5)x3</summary>
```{r echo=FALSE}
4+5*3>(4+5)*3

```

<div style="margin-bottom: 1000px;">
</div>


# Variables

Variables hold objects that are assigned to them. 

```{r}
a <- 1
```

They can be identical to the object assigned to them

```{r}
a <- 1
b <- a

#does b equal 1?

b==1
```

Variables enable complex operations on data

```{r}
h <- 2^100
i <- h/3E100
j <- 1E5
k <- j^(-1*i)

#is k greater than 1?

k < 1
```

<div style="margin-bottom: 1000px;">
</div>

# Complex Data Structures

## Vectors

A Vector is an ordered collection of either numerics, characters, or logicals

```{r}
c(1, 2, 3)
c(TRUE, FALSE, TRUE)
c("hello", "world")
```

<div style="margin-bottom: 1000px;">
</div>

### Challenge Question 5

A vector in R has a limitation: they can only allow one type of class in each vector.

If you add more than one type of class, it will default to a character. 

With that said, what would the following give you? 

```{r echo=FALSE}

class(c("puppies", 2))

```

<details>
<summary>What is the class of (c("puppies", 2))</summary>
```{r}
class(c("puppies", 2))
```
</details>
<div style="margin-bottom: 1000px;">
</div>

### Challenge Question 6

I can change the following vector in my variable to all numbers

```{r}
my_num <- as.numeric(c("2", 2))

class(my_num)
```

But what would happen if I tried to do the following? 

```{r warning=FALSE, results='hide'}
as.numeric(c("salt", "2"))

print(as.numeric(c("salt", "2")))

```

<details>
<summary>What is the class of as.numeric(c("salt", "2"))</summary>

```{r error=TRUE, echo=FALSE}

print(as.numeric(c("salt", "2")))

class(as.numeric(c("salt", "2")))
```
</details>

<div style="margin-bottom: 1000px;">
</div>

# Vectors Continued

Vectors can also have a vector of names which describe each element

```{r}
grades <- c(98, 95, 82)
names(grades) <- c("Jimmy", "Alice", "Susan")

grades

```

Elements from a vector can be accessed using the index of the desired data. Lets say I wanted to know what Alice's grade was.

```{r}
grades[2]
```


But lets say that I didn't know Alice was the second index

```{r}
grades["Alice"]
```

If you want to add a series of numbers to a vector of integers, you can do the following: 

```{r}
my_ints <- 1:10
my_ints
```

<div style="margin-bottom: 1000px;">
</div>

# Lists

A list is an ordered collection of any objects. This differs from vectors, because vectors can only be one type of object class

```{r}
my_list <- list(1, "b", TRUE, c(1,2,3))
my_list
```

Lists can also have names

```{r}
names(my_list)=c("My #", "My Letter", "My logical", "My_num_Set")

my_list
```

The elements within a list can be accessed by using numeric indexes or by the element name

```{r}
my_list[[2]]

my_list[["My_num_Set"]]

my_list$My_num_Set

#If I did the following, I would get the whole vector within the list

my_list[2]

```

<div style="margin-bottom: 1000px;">
</div>

#Data Frames

Data frames are similar to excel sheets. They are 2D arrays which can hold numeric, character, and boolean data. They also have column names. 

```{r}
my_df <- data.frame("students"=c("Jimmy", "Alice", "Susan"), 
                    "Grades"=c(98, 95, 82))


my_df
```

Data frames can be accessed numerically by expressing the row and column of interest. 

```{r}
#What grade did Susan get? 

my_df[3,2] #row, column

```

They can also be accessed with the $ sign

```{r}
my_df$Grades

#What grade did Alice get?

my_df$Grades[2]

```


<div style="margin-bottom: 1000px;">
</div>


# End of the workshop

Thank you!

Questions?

<div style="margin-bottom: 500px;">
</div>















